---
title: "Armour change analysis"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

```{r}
library(phytools)
library(dispRity)
library(Inapp) # home package
library(landmarktest) # home package
source("../Functions/clade.change_fun.R")
source("../Functions/utilities.R")
set.seed(42)

## Loading the extant data
load("../Data/Processed/extant.rda")
```

Here we want to test:

  1. Whether there is a difference in body size between armoured/non-armoured fishes
  2. Whether there is a difference in body size at the time armour appears/disappear in a phylogeny
  3. Whether the there is a relation between the body size or armoured fishes and the predator's size

If all three are true, we can suggest that armour evolved in fishes as a response to predation.

# Is a difference in body size between armoured/non-armoured fishes?

This is a crude analysis simply to test if there is a difference between the fishes with and without armour.
A simple t-test can be used here.

```{r}
## The tree
tree <- spine_data$tree

## The body size values
body_size_values <- spine_data$spine_data[, 3]

## Armour values
armour_values <- spine_data$armour[,1]

## Selecting both groups
armour_body_sizes <- spine_data$spine_data[which(spine_data$armour[,1] == 1), 3]
nonarm_body_sizes <- spine_data$spine_data[which(spine_data$armour[,1] == 0), 3]

## Logging the data
armour_body_sizes <- log(armour_body_sizes)
nonarm_body_sizes <- log(nonarm_body_sizes)

## Testing the differences
armour_bs_test <- t.test(armour_body_sizes, nonarm_body_sizes)

## Plotting this
boxplot(list("Spines" = armour_body_sizes, "No spines" = nonarm_body_sizes),
        ylab = "Log(body size)")
text(1, -0.25, paste0("Mean difference = ",
  round(abs(armour_bs_test$estimate[1]-armour_bs_test$estimate[2]), digits = 3),
  "\np = ", round(armour_bs_test$p.value, digits = 3)))
```

Nice.
Something is happening.

## Correlation between body length and armourness

We can also measure the correlation between body length and armourness more precisely using the Threshold model (`phytools::threshBayes`).

```{r, fig.height = 6, fig.width = 6}
## Creating the matrix of traits
trait_matrix <- cbind(spine_data$armour[, 1], body_size_values)

## Threshold model
#mcm_binary_1M <- threshBayes(tree, trait_matrix, types = c("discrete","continuous"), ngen = 1000000)
#save(mcm_binary_1M, file = "../Data/Processed/MCMC_living_binary_1M.rda")
load("../Data/Processed/MCMC_living_binary_1M.rda")

## The MCMC results
mcm_binary_1M
plot(mcm_binary_1M)
```

Excellent: there is a negative correlation between body length and armourness: the smaller you get, the more likely you are to evolve have spines!

This is how the the phylogeny and the reconstructed states look:


```{r, fig.height = 10, fig.width = 10}
## get the node changes
results_topol <- run.timespines(body_size_values, armour_values, tree,
                                method = "topol",
                                taxa.names = rownames(spine_data$spine_data))

## Create a vector of two colours
spine_cols <- c("white", "black")

## Calculating the traits ancestral values
traits <- body_size_values[match(tree$tip.label, rownames(spine_data$spine_data))]
names(traits) <- tree$tip.label
trait_map <- contMap(tree, traits, plot = FALSE)

## Results plot
par(mar = c(1,1,1,1))
plot(trait_map, fsize = 0.001, type = "fan", lwd = 2)

## Adding the spine colours
tiplabels(pch = 21, cex = 0.7, col = "black")
tiplabels(pch = 16, cex = 0.7,
          col = spine_cols[armour_values[match(tree$tip.label, rownames(spine_data$spine_data))]+1])

## Adding the node changes
nodelabels(pch = 21, node = unlist(results_topol$node.changes), col = "black", bg = "magenta",
           cex = 1.2)

## Legends
legend("topleft",legend = c("Not spiny", "Spiny", "Stage change"), pch = 21, col = "black",
       pt.bg = c(spine_cols, "magenta"), bty = "n")
legend("bottomleft",legend = paste(round(trait_map$lims, digit = 2), collapse = " Body length "),
       bty = "n")

```

Now this is just a reconstruction of how the two traits (body size and armourness) coevolved but it doesn't allow to really distinguish whether there is an effect of body size when the armourness trait changes: this difference only tells us armoured fishes are smaller than the others.

# Is a difference in body size at the time armour appears/disappear in a phylogeny

So now lets see whether the changes in armourness occurs when there is a change in body size.
The hypothesis being that:

 1. Is there a relation between change in armourness regime (i.e. presence or absence) and body length within clades.
 In other words, are taxa with/without armour bigger/smaller than the rest.

 2. If 1 is true, is there a relation between members in a clade having armour and the size of the top predator.

To check these hypothesis, we use two approaches based on topology only or on topology and branch length.
Both approach are based on the same principles:

  1. Estimating the armour origination state of a clade (i.e. the state of armourness at the appearance of the clade)
  2. Measure the body length of taxa in the clade that do not have the origination state.

For the second hypothesis, we can then make these body lengths relative to the biggest predator's length.

The topology and topology+branch length method only differ in the way to measure the armourness origination state:

 1. When based on topology only, the origination state is the state that is represented in majority in the clade (parsimony-like --- if there is no strict majority, the origination state is arbitrarily set to no-armour (0))
 2. When based on topology and branch length, the origination state is estimated using a likelihood ancestral state estimation calculation based on the taxa' states and branch length
 + The length at the node is used.

See analysis 02 for discussion on the pros and cons of both methods.

```{r, fig.width = 6, fig.height = 6}
## Running the results for topology
results_topol <- run.timespines(body_size_values, armour_values, tree,
                                method = "topol",
                                taxa.names = rownames(spine_data$spine_data))

## Running the results for topology and branch length
results_brlen <- run.timespines(body_size_values, armour_values, tree,
                                method = "brlen",
                                taxa.names = rownames(spine_data$spine_data))
```

We can now visualise where do the group changes occur relative to their group size:

```{r, fig.width = 6, fig.height = 12}
par(mfrow = c(2,1))
## Plotting the pooled data for every group
plot.change(results_topol$normal.val, results_topol$change.val, 
            main = "Topology", xlab = "Scaled log BL")
## Plotting the pooled data for every group
plot.change(results_brlen$normal.val, results_brlen$change.val,
            main = "Topology and branch length", xlab = "Scaled log BL")
```

We then need to test whether this difference is significant or not.

## Testing the body size at time of appearance

For that we can use a kind of permutation test (here an exact bootstrap test).
This checks whether a statistic observed in a subset of a distribution (here the median) is different from the same statistic measured in the overall distribution.

If the results are out of the range from the statistics from the overall distribution ($p < 0.05$) then there is a different between the subset distribution (body size at time of appearance) and the rest of the distribution.

The values compared are $median_bs - median_obs$, positive values indicate that the observed data is smaller

```{r}
## Running the test for the topology
## Adding the estimate ancestral values to the pooled distribution
distribution_topol <- c(unlist(results_topol$normal.val),
                        unlist(results_topol$change.val))

## The IDs to test are the last one in the distribution
subset_topol <- seq(from = length(unlist(results_topol$normal.val)) + 1,
                    to = length(distribution_topol))

## Applying the random test for the topology
test_topol <- bootstrap.test(distribution_topol, subset_topol, statistic = median, replicates = 500,
                             rarefaction = TRUE)

## Running the test for the branch length
## Adding the estimate ancestral values to the pooled distribution
distribution_brlen <- c(unlist(results_brlen$normal.val),
                        unlist(results_brlen$change.val))

## The IDs to test are the last one in the distribution
subset_brlen <- seq(from = length(unlist(results_brlen$normal.val)) + 1,
                    to = length(distribution_brlen))

## Applying the random test for the branch length
test_brlen <- bootstrap.test(distribution_brlen, subset_brlen, statistic = median, replicates = 500,
                             rarefaction = TRUE)
```

```{r, fig.width = 6, fig.height = 12}
par(mfrow = c(2,1))
## Plotting the pooled data for every group
plot(test_topol, main = "Topology", xlab = "Median difference")
add.rare.plot(test_topol)
## Add the text
legend(ifelse(test_topol$obs < test_topol$expvar[2], "topleft", "topright"),
  paste0("p = ", round(test_topol$pvalue, 5), sep = "\n")  , bty = "n")


## Plotting the pooled data for every group
plot(test_brlen, main = "Branch length", xlab = "Median difference")
add.rare.plot(test_brlen)
## Add the text
legend(ifelse(test_brlen$obs < test_brlen$expvar[2], "topleft", "topright"),
  paste0("p = ", round(test_brlen$pvalue, 5), sep = "\n")  , bty = "n")
```

These bootstraps tests show that the difference in median between the body length at the change in armourness is significantly smaller than the global median body lengths.
This is true even when taking into account the number of events of changes in armourness (rarefaction).

## Testing the whether this difference is linked to the biggest predator size

Here we can perform exactly the same test but by scaling the results

```{r}
## Scaling the distributions
sca_distribution_topol <- distribution_topol/max(distribution_topol)
sca_distribution_brlen <- distribution_brlen/max(distribution_brlen)

## Applying the random tests
sca_test_topol <- bootstrap.test(sca_distribution_topol, subset_topol, statistic = median,
                                 replicates = 500, rarefaction = TRUE)

## Applying the random test for the branch length
sca_test_brlen <- bootstrap.test(sca_distribution_brlen, subset_brlen, statistic = median,
                                 replicates = 500, rarefaction = TRUE)
```

```{r, fig.width = 6, fig.height = 12}
par(mfrow = c(2,1))
## Plotting the pooled data for every group
plot(sca_test_topol, main = "Topology (scaled)", xlab = "Median difference")
add.rare.plot(sca_test_topol)
## Add the text
legend(ifelse(sca_test_topol$obs < sca_test_topol$expvar[2], "topleft", "topright"),
  paste0("p = ", round(sca_test_topol$pvalue, 5), sep = "\n")  , bty = "n")


## Plotting the pooled data for every group
plot(sca_test_brlen, main = "Branch length (scaled)", xlab = "Median difference")
add.rare.plot(sca_test_brlen)
## Add the text
legend(ifelse(sca_test_brlen$obs < sca_test_brlen$expvar[2], "topleft", "topright"),
  paste0("p = ", round(sca_test_brlen$pvalue, 5), sep = "\n")  , bty = "n")
```

Unsurprisingly the results are not different.

> Need to improve this test. Not sure if it's savy.

> Need to correct the p-values.