---
title: "Armour change analysis"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

# Pipeline for looking at (dis)appearance of armour per clade in extant clades


```{r}
library(dispRity)
library(Inapp)
source("../Functions/clade.change_fun.R")
set.seed(42)

## Loading the extant data
load("../Data/Processed/extant.rda")
```

Here we test two hypothesis:

 1. Is there a relation between change in armourness regime (i.e. presence or absence) and body length within clades.
 In other words, are taxa with/without armour bigger/smaller than the rest.

 2. If 1 is true, is there a relation between members in a clade having armour and the size of the top predator.

To check these hypothesis, we use two approaches based on topology only or on topology and branch length.
Both approach are based on the same principles:

  1. Estimating the armour origination state of a clade (i.e. the state of armourness at the appearance of the clade)
  2. Measure the body length of taxa in the clade that do not have the origination state.

For the second hypothesis, we can then make these body lengths relative to the biggest predator's length.

The topology and topology+branch length method only differ in the way to measure the armourness origination state:

 1. When based on topology only, the origination state is the state that is represented in majority in the clade (parsimony-like --- if there is no strict majority, the origination state is arbitrarily set to no-armour (0))
 2. When based on topology and branch length, the origination state is estimated using a likelihood ancestral state estimation calculation based on the taxa' states and branch length
 + The mass at the node is used.

See analysis 02 for discussion on the pros and cons of both methods.

## Relation between change in armourness regime and body length

### Topology method


```{r}
## The tree
tree <- spine_data$tree

## The body size values
body_size_values <- spine_data$spine_data[, 3]

## The armour values
armour_binary <- paste(spine_data$armour[,1], collapse = "") # character_rand
# armour_scored <- paste(spine_data$armour[,2], collapse = "") # character_rand

## Traversal of the tree to detect changes.
traversal_binary <- apply.reconstruction(tree, armour_binary, passes = 1,
                                         method = "Fitch", match.tip.char = TRUE)

## Get the subtrees with a state change
subtrees_binary <- sapply(traversal_binary$changes,
                         function(X, tree) return(extract.clade(tree, node = X)),
                         tree = tree, simplify = FALSE)

## Get the changes per clades
changes_binary <- lapply(subtrees_binary, topology.change, data = spine_data$spine_data,
                         state = spine_data$armour[,1], data.out = 3)

## Extract the changes values
binary_normal <- lapply(extract.change(changes_binary, what = "group.val"), log)
binary_change <- lapply(extract.change(changes_binary, what = "changed.val"), log)

## get the maximum of each group
max_normal <- lapply(binary_normal, max, na.rm = TRUE)

## Normalising the body sizes relative to their group maximum
scale.fun <- function(X, Y) scale(X, center = FALSE, scale = Y)
binary_normal_scaled <- mapply(scale.fun, binary_normal, max_normal, SIMPLIFY = FALSE)
binary_change_scaled <- mapply(scale.fun, binary_change, max_normal, SIMPLIFY = FALSE)
```

We can now visualise where do the group changes occur relative to their group size

```{r, fig.width = 12, fig.height = 6}
## Plotting the pooled data for every group
plot.change(binary_normal_scaled, binary_change_scaled, main = "Topology - Binary change", xlab = "Scaled log BL")
```


### Topology + branch length method


```{r}
## Performing the ancestral states reconstruction
ace_body_size <- ace(body_size_values[match(tree$tip.label, rownames(spine_data$spine_data))],
                     tree, method = "pic")

## Extract the changes for each group
ace_changes <- log(ace_body_size$ace[traversal_binary$changes-Ntip(tree)])

## Scale these values using the maximums from option 2
ace_changes_scaled <- mapply(scale.fun, as.list(ace_changes), max_normal, SIMPLIFY = FALSE)
```

> Note that the method for the reconstructions is important here! I just go with the simple PIC method for convenience

Similarly we can now plot the differences:

```{r, fig.width = 12, fig.height = 6}
## Plotting the pooled data for every group
plot.change(binary_normal_scaled, ace_changes_scaled, main = "Branch length - Random",
            xlab = "Scaled log BS")

```
